MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-07-11 main/MSstatsPTM_Main.tex
updated protein wording

This manuscript proposes a versatile statistical analysis framework that accurately detects relative changes in post-translational modifications. The framework requires an experimental design, which quantifies, for each sample, both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites. The framework supports data-dependent acquisitions (DDA) that are label-free or tandem mass tag (TMT)-based. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design. Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources.

We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations. The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations. By appropriately leveraging the information from the unmodified peptides, the proposed approach improved the accuracy of the estimates of PTM fold changes, and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods. In particular, accounting for the confounding from protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.

@@

Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition. For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate corresponds to an LC-MS/MS run.  
Figure \ref{fig:data-structure} schematically illustrates this data structure for one PTM site, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
For one protein, the PTM site is represented by $K$ spectral features (i.e., peptide ions, distinguished by their cleavage residues and charge states). 
The number of modified and unmodified features typically varies across proteins. Some log-intensities may be outliers, and some spectral features can be missing. 
The log$_2$-intensity of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.

@@

The basic two-sample $t$-test~\cite{kutner_etal_04a} is based on the model
\begin{eqnarray}
y^*_{ijk} = \hat{\mu}^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}$ and its standard error.
Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=2J-2$ degrees of freedom in balanced designs.
Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the protein.

@@

\todo{DK: Current plan is to follow what Ting did for MSstatsTMT and keep this part brief but expand on EB methods in the supplementary.}\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}. In particular, the ratio-based version which models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}. Limma uses the transformed inputs to fit a linear model with a fixed condition and run effect for each PTM.

@@

We leverage the existing methods in $MSstats$ and $MSstatsTMT$ and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.  Each feature type is summarized separately using $MSstats$ or $MSstatsTMT$, resulting in run level values for the PTM, $\hat{y}_{ij}^{\ast}$, and unmodified peptide. The estimated summaries of the modified features are then used as the input to the model, such as in \eqref{msstats_model}, in $MSstats$ or $MSstatsTMT$. Similarly, the estimated summaries of the unmodified features are used as the input to the corresponding model in $MSstats$ or $MSstatsTMT$.

@@

\todo{Could we try to use a consistent vocabulary for 'global protein' (we sometimes use unmodified peptides, unmodified features, global profiling etc.) Could you check the entire manuscript and see that we use the same terminology for that}

The proposed methods are implemented in the open source R package $MSstatsPTM$, overviewed in~\figref{msstatsptm_design}. \todo{$MSstatsPTM$ is applicable to most experimental designs, label-free data acquisitions such as XXX and label-based data acquisitions such as XXX} $MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.  \todo{Could you rephrase the next sentence to be more specific. If someone is a tool developer, what exactly should they do?} To identify the local modification site we require the tool to identify where in the sequence the modification occurred.  The package includes converters for seamless integration with multiple such tools. \todo{Could you clarify what the converters do, especially for PTM? Do we discard peptides with multiple modification sites?}. For experimental designs that include separate runs enriched in modified peptides and global profiling runs,  conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs. For experimental designs that include modified and unmodified peptides in a same run, \todo{what happens for the conversion?}

\todo{Could you rephrase the next sentence? What is summarized on the PTM side? The peptide sequence? Or any peptide sequence that contains the modification site?} The next step is peptide/protein summarization. Summarization is performed separately for the PTM and global protein in the same way as described above for $MSstats$. When summarizing the unmodified peptide features, the package optionally imputes missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}. When summarizing the modified features, missing value imputation is also possible but should be performed with care. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can violate the assumptions underlying the imputation, and lead to numerically unstable  results. Next, separate statistical models are fit to both the summaries of the modified features and the to summaries of the unmodified features. The models are automatically selected to reflect the experimental design and the data acquisition. If the base model is not applicable for a particular PTM or protein, e.g. due to missing data, a simplified model is fit.

@@

We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTMs. The true positives were defined as PTMs with changes distinct from the overall changes in protein abundance. The true negatives were defined as PTMs which, after accounting for the changes in the overall protein abundance, were not differentially abundant. Additional summaries were the false discovery rate $FDR=FP/(TP+FP)$, $Recall=TP/(TP + FN)$, and $Accuracy=(TP + TN) / (TP + TN + FP + FN)$.
For biological experiments with unknown ground truth, we compared the differentially abundant PTMs  with and without adjusting for changes in unmodified protein abundance.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
information \subsubsection*{In the label-free benchmark experiment, MSstatsPTM had a higher sensitivity than the existing methods}

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than $Limma$ and 32.98\% smaller than $t$-test's. While Limma and $t$-test generally estimated the correct fold change, the proposed method's estimation was more consistent across all modifications.

@@

The results of these experiments are summarized in~\figref{venn_diagrams} and~\figref{profile_plots}. \figref{data4_venn_diagram} and~\figref{data5_venn_diagram} show the number of differentially abundant modified peptides before and after adjustment. Adjusting for changes in the global protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa. A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became statistically insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When the test was applied to Dataset 4, only one peptide became non-differentially abundant due to an increase in variance. When applied to Dataset 5, 548 peptides became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became became non-differentially abundant). Thus we can conclude that the drop off in differentially abundant peptides was due to changes in global protein abundance, not simply increased model variance.

Interestingly, the test above showed that the global protein was more likely to cause observed changes in PTM abundance, rather than mask changes in the PTM. The later case occurs when the global protein exhibits a change in abundance, while the modification does not. An example of this can be seen in \figref{data5_profile_plot}. Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$. However the global protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment. This caused the modification to go from differentially abundant to non-differentially abundant. As discussed previously, these cases occur frequently and can be thought of as false positives that would occur without accounting for confounding with the global protein. The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.

Conversely, cases where the global protein masks changes in the PTM are more infrequent. In this case, the modification and protein exhibit contradictory changes in abundance, moving in opposite directions. One such case can be seen in~\figref{data4_profile_plot}. Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by dox treatment. Our reanalysis confirmed this, with the global $GSDMD$ protein being downregulated when Dox treatments reached the 4 and 6 hour marks. Conversely, ubiquitination of $GSDMD$ at site $K62$   upregulated abundance between the same conditions. This upregulation was originally confounded by the downregulation of unmodified $GSDMD$, appearing to have little change between no dox and dox 4 and 6 hour. The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox1hr and Dox4hr conditions. This was an interesting result, as the modification contradicts what previous research showed for the global protein. This effect can be thought of as a false negative and would have been more challenging to observe without the proposed approach. There are many non-differentially abundant PTMs, and having to comb through each one individually to determine false negatives is generally an unfeasible task. The proposed approach removes the confounding on an experimental level, revealing interesting PTMs that are entirely masked due to changes in the global protein.

@@

As discussed in \secref{exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment. Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 ubiquitin sites identified, only 4,526 had a corresponding unmodified protein and could be adjusted. Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model compared to other experiments, seen in Table \ref{tab:dataDescription}.

The number of differentially abundant PTMs before and after adjustment are shown in~\figref{data6_plots}. After adjusting for changes in the unmodified protein, there were fewer significant modified peptides. However, this was mainly due to not having a global profiling run. When looking only at PTMs with a corresponding global protein, the number of differentially modified PTMs exhibited a small drop, but was more consistent. As above, we tested if this drop in differentially abundant PTMs was due to an increase in variance. Here only 25 PTMs became insignificant due to increased variance. When there was no global profiling run the methods were less effective at removing confounding with the global protein. This was entirely due to not being able to perform adjustment without a corresponding global protein. In this case adjustment caused us to toss out a large number of the PTMs in the original model. In cases without a global profiling run, the proposed method can still be used, but it should be used in conjunction with the unadjusted model so that information is not lost.

>>>>

This manuscript proposes a versatile statistical analysis framework that accurately detects relative changes in post-translational modifications. The framework requires an experimental design, which quantifies, for each sample, both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites. The framework supports data-dependent acquisitions (DDA) that are label-free or tandem mass tag (TMT)-based. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design. Next, model-based inferences regarding the PTM and the unmodified protein-level abundances are combined to account for the confounding between these two sources.

We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations. The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations. By appropriately leveraging the information from the unmodified proteins, the proposed approach improved the accuracy of the estimates of PTM fold changes, and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods. In particular, accounting for the confounding from unmodified protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.

@@

Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition. For simplicity, we assume that the experiment has no technical replicates, such that each biological replicate corresponds to an LC-MS/MS run.  
Figure \ref{fig:data-structure} schematically illustrates this data structure for one PTM site, in a special case of $I=2$ conditions and $J=2$ biological replicates per condition.
For one protein, the PTM site is represented by $K$ spectral features (i.e., peptide ions, distinguished by their cleavage residues and charge states). 
The number of modified and unmodified features typically varies across proteins. Some log-intensities may be outliers, and some spectral features can be missing. 
The log$_2$-intensity of Feature $k$, in Replicate $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified protein in the same run is denoted by $y_{ijl}$. The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.

@@

The basic two-sample $t$-test~\cite{kutner_etal_04a} is based on the model
\begin{eqnarray}
y^*_{ijk} = \hat{\mu}^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}$ and its standard error.
Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=2J-2$ degrees of freedom in balanced designs.
Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.

@@

\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}. In particular, the ratio-based version which models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}. Limma uses the transformed inputs to fit a linear model with a fixed condition and run effect for each PTM.

@@

We leverage the existing methods in $MSstats$ and $MSstatsTMT$ and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.  Each feature type is summarized separately using $MSstats$ or $MSstatsTMT$, resulting in run level values for the PTM, $\hat{y}_{ij}^{\ast}$, and unmodified protein. The estimated summaries of the modified features are then used as the input to the model, such as in \eqref{msstats_model}, in $MSstats$ or $MSstatsTMT$. Similarly, the estimated summaries of the unmodified features are used as the input to the corresponding model in $MSstats$ or $MSstatsTMT$.

@@

The proposed methods are implemented in the open source R package $MSstatsPTM$, overviewed in~\figref{msstatsptm_design}. $MSstatsPTM$ is applicable to most experimental designs, label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT. $MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.  \todo{Could you rephrase the next sentence to be more specific. If someone is a tool developer, what exactly should they do?} To identify the local modification site we require the tool to identify where in the sequence the modification occurred.  The package includes converters for seamless integration with multiple such tools. \todo{Could you clarify what the converters do, especially for PTM? Do we discard peptides with multiple modification sites?}. For experimental designs that include separate runs enriched in modified peptides and global profiling runs,  conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs. For experimental designs that include modified and unmodified peptides in a same run, \todo{what happens for the conversion?}

The next step is PTM/protein summarization. Summarization is performed separately for the PTM and unmodified protein. When summarizing the PTM, modified peptide features that span the same modification site are summarized together. Peptides that include multiple modifications are not included in the single modification summarization, and are grouped separately. The unmodified protein summarization is performed as discussed above for $MSstats$. When summarizing the unmodified protein features, the package optionally imputes missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}. When summarizing the modified features, missing value imputation is also possible but should be performed with care. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can violate the assumptions underlying the imputation, and lead to numerically unstable  results. Next, separate statistical models are fit to both the summaries of the modified features and the to summaries of the unmodified features. The models are automatically selected to reflect the experimental design and the data acquisition. If the base model is not applicable for a particular PTM or protein, e.g. due to missing data, a simplified model is fit.

@@

We evaluated $MSstatsPTM$ on simulated and spike-in datasets with known ground truth in terms of true positives ($TP$), false positives ($FP$), true negatives ($TN$), and false negatives ($FN$) differentially abundant PTMs. The true positives were defined as PTMs with changes distinct from the overall changes in unmodified protein abundance. The true negatives were defined as PTMs which, after accounting for the changes in the overall protein abundance, were not differentially abundant. Additional summaries were the false discovery rate $FDR=FP/(TP+FP)$, $Recall=TP/(TP + FN)$, and $Accuracy=(TP + TN) / (TP + TN + FP + FN)$.
For biological experiments with unknown ground truth, we compared the differentially abundant PTMs with and without adjusting for changes in unmodified protein abundance.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In the label-free benchmark experiment, MSstatsPTM had a higher sensitivity than the existing methods}

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in unmodified protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (\figref{spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than $Limma$ and 32.98\% smaller than $t$-test's. While Limma and $t$-test generally estimated the correct fold change, the proposed method's estimation was more consistent across all modifications.

@@

The results of these experiments are summarized in~\figref{venn_diagrams} and~\figref{profile_plots}. \figref{data4_venn_diagram} and~\figref{data5_venn_diagram} show the number of differentially abundant modified peptides before and after adjustment. Adjusting for changes in the unmodified protein caused more PTMs to change from differentially abundant to non-differentially abundant than vice versa. A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became statistically insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When the test was applied to Dataset 4, only one PTM became non-differentially abundant due to an increase in variance. When applied to Dataset 5, 548 PTMs became non-differentially abundant due to an increase in variance (only 3.4\% of all PTMs that became became non-differentially abundant). Thus we can conclude that the drop off in differentially abundant PTMs was due to changes in global protein abundance, not simply increased model variance.

Interestingly, the test above showed that the unmodified protein was more likely to cause observed changes in PTM abundance, rather than mask changes in the PTM. The later case occurs when the unmodified protein exhibits a change in abundance, while the modification does not. An example of this can be seen in \figref{data5_profile_plot}. Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$. However the unmodified protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment. This caused the modification to go from differentially abundant to non-differentially abundant. As discussed previously, these cases occur frequently and can be thought of as false positives which are caused by confounding with the unmodified protein. The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.

Conversely, cases where the unmodified protein masks changes in the PTM are more infrequent. In this case, the modification and unmodified protein exhibit contradictory changes in abundance, moving in opposite directions. One such case can be seen in~\figref{data4_profile_plot}. Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by dox treatment. Our reanalysis confirmed this, with the $GSDMD$ protein being downregulated when Dox treatments reached the 4 and 6 hour marks. Conversely, ubiquitination of $GSDMD$ at site $K62$ upregulated abundance between the same conditions. This upregulation was originally confounded by the downregulation of unmodified $GSDMD$, appearing to have little change between no Dox and Dox 4 and 6 hour. The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox 1 hour and Dox 4 hour conditions. This was an interesting result, as the modification contradicts what previous research showed for the unmodified protein. This effect can be thought of as a false negative and would have been challenging to observe without the proposed approach. There are many non-differentially abundant PTMs, and having to comb through each one individually to determine false negatives is generally an unfeasible task. The proposed approach removes the confounding on an experimental level, revealing interesting PTMs that are entirely masked due to changes in the unmodified protein.

@@

As discussed in \secref{exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment. Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a lack of overlap between PTMs and unmodified proteins. Any PTM without a corresponding unmodified protein could not be adjusted. Of the 10,799 ubiquitin sites identified, only 4,526 had a corresponding unmodified protein and could be adjusted. Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model compared to other experiments, seen in Table \ref{tab:dataDescription}.

The number of differentially abundant PTMs before and after adjustment are shown in~\figref{data6_plots}. After adjusting for changes in the unmodified protein, there were fewer differentially modified peptides. However, this was mainly due to not having a global profiling run. When looking only at PTMs with a corresponding unmodified protein, the number of differentially modified PTMs exhibited a small drop, but was more consistent. As above, we tested if this drop in differentially abundant PTMs was due to an increase in variance. Here only 25 PTMs became insignificant due to increased variance. When there was no global profiling run the methods were less effective at removing confounding with the unmodified protein. This was entirely due to not being able to perform adjustment without a corresponding global protein. In this case adjustment caused us to toss out a large number of the PTMs in the original model. In cases without a global profiling run, the proposed method can still be used, but it should be used in conjunction with the unadjusted model so that information is not lost.
